---
title: "EDA"
format: html
editor: visual
---

## Introduction

The data set we are working with is the Diabetes Health Indicators Dataset. The data set contains 22 variables for 253,680 individuals. Those variables include whether or not a person has diabetes, demographic factors, and lifestyle characteristics. Essentially, we will be using the variables in this data set to create a model to predict whether or not a person has diabetes.

In this analysis, the important variables we are considering are HighBP, HighChol, BMI, AnyHealthcare, Age, GenHlth, and Income. HighBP and HighChol are binary variables for whether or not a person has high blood pressure or high cholesterol, respectively. BMI is the body mass index. AnyHealthcare is a binary variable for having any type of health care access. Age places an individual's age into one of 13 different age groups. GenHlth has an individual rate their overall health from 1-5. Income places an individual's income into one of 8 different groups.

Here we will perform some exploratory data analysis, do gain a better understanding of our data. 

## EDA

```{r}
#| include: false
#| echo: false
library(tidyverse)
library(ggcorrplot)
```

```{r}
my_diabetes_data<-read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")
```

First, we check for missing data. There is none.

```{r}
colSums(is.na(my_diabetes_data))
unique(my_diabetes_data$Diabetes_binary)
```

We'll convert many of our variables into factors with meaningful names. 

```{r}
my__data<-my_diabetes_data|>
  mutate(Fruits=as.factor(ifelse(Fruits==0, "No fruit", "Fruit")),
         Stroke=as.factor(ifelse(Stroke==0, "No stroke", "Stroke")),
         Smoker=as.factor(ifelse(Smoker==0, "No 100 cigs", "100 cigs")),
         HeartDiseaseorAttack=as.factor(ifelse(HeartDiseaseorAttack==0, "No CHD/MI", "CHD/MI")),
         PhysActivity=as.factor(ifelse(PhysActivity==0, "No physical activity past 30 days", "Physical activity past 30 days")),
         CholCheck=as.factor(ifelse(CholCheck==0, "No cholesterol check in 5 years", "Cholesterol check in past 5 years")),
         HighChol=as.factor(ifelse(HighChol==0, "No high cholesterol", "High cholesterol")),
         HighBP=as.factor(ifelse(HighBP==0, "No high blood pressure", "High blood pressure")),
         Diabetes_binary=ifelse(Diabetes_binary==0, "No diabetes", "Diabetes"),
         Veggies=as.factor(ifelse(Veggies==0, "No veggies", "Veggies")),
         HvyAlcoholConsump=as.factor(ifelse(HvyAlcoholConsump==0, "Not heavy Drinker", "Heavy Drinker")),
         AnyHealthcare=as.factor(ifelse(AnyHealthcare==0, "No healthcare", "Healthcare")),
         NoDocbcCost=as.factor(ifelse(NoDocbcCost==0, "No doctor", "Doctor")),
         GenHlth=as.factor(ifelse(GenHlth==1, "Excellent", ifelse(GenHlth==2, "Very good", ifelse(GenHlth==3, "Good", ifelse(GenHlth==4, "Fair", "Poor"))))),
         DiffWalk=as.factor(ifelse(DiffWalk==0, "Can walk", "Difficult walking")),
         Sex=as.factor(ifelse(Sex==0, "female", "male")),
         Age=as.factor(Age),
         Education=as.factor(Education),
         Income=as.factor(Income)
         )

```

Next, I'm going to create a correlation matrix, to understand the realtionship between the variables. This matrix helps to decide which variables will be important for our model.

```{r}
corr <- cor(my_diabetes_data)
ggcorrplot(corr,
           type = "lower")
```

Create bar plots or histograms for the variables we care about.

```{r}
ggplot(my__data, aes(x=BMI)) + 
  geom_histogram()

ggplot(my__data, aes(x=HighBP)) + 
  geom_bar()

ggplot(my__data, aes(x=HighChol)) + 
  geom_bar()

ggplot(my__data, aes(x=AnyHealthcare)) + 
  geom_bar()

ggplot(my__data, aes(x=Age)) + 
  geom_bar()

ggplot(my__data, aes(x=GenHlth)) + 
  geom_bar()

ggplot(my__data, aes(x=Income)) + 
  geom_bar()

ggplot(my__data, aes(x=Diabetes_binary)) + 
  geom_bar() 
```

This bar plot and table show the relationship between Diabetes and General Health, putting emphasis on the presence of diabetes in the different health categories.

```{r}
table(my__data$Diabetes_binary, my__data$GenHlth)
ggplot(my__data, aes(x=Diabetes_binary))+
  geom_bar()+
  facet_wrap(~GenHlth)
```

Our correlation matrix shows a relationship between General Health and Income. We'll create an additional bar plot that shows that the distribution of General Health is not consistent across income levels.

```{r}
table(my__data$GenHlth, my__data$Income)
ggplot(my__data, aes(x=GenHlth))+
  geom_bar()+
  facet_wrap(~Income)
```

Just income and diabetes.

```{r}
table(my__data$Diabetes_binary, my__data$Income)
ggplot(my__data, aes(x=Diabetes_binary))+
  geom_bar()+
  facet_wrap(~Income)
```

Distribution of diabetes based on blood pressure level.

```{r}
table(my__data$Diabetes_binary, my__data$HighBP)
ggplot(my__data, aes(x=Diabetes_binary))+
  geom_bar()+
  facet_wrap(~HighBP)
```

Distribution of diabetes based on cholesterol level.

```{r}
table(my__data$Diabetes_binary, my__data$HighChol)
ggplot(my__data, aes(x=Diabetes_binary))+
  geom_bar()+
  facet_wrap(~HighChol)
```

Distribution of diabetes based on healthcare access.

```{r}
table(my__data$Diabetes_binary, my__data$AnyHealthcare)
ggplot(my__data, aes(x=Diabetes_binary))+
  geom_bar()+
  facet_wrap(~AnyHealthcare)
```

Distribution of diabetes based on age.

```{r}
table(my__data$Diabetes_binary, my__data$Age)
ggplot(my__data, aes(x=Diabetes_binary))+
  geom_bar()+
  facet_wrap(~Age)
```

Summary statistics and histogram of BMI for each diabetes classification.

```{r}
my__data |>
  summarize(mean=mean(BMI), sd=sd(BMI), range = (max(BMI)-min(BMI)))
my__data |>
  group_by(Diabetes_binary)|>
  summarize(mean=mean(BMI), sd=sd(BMI), range = (max(BMI)-min(BMI)))
ggplot(my__data, aes(x=BMI))+
  geom_histogram()+
  facet_wrap(~Diabetes_binary)
```